---
title: "What is the enrollment distribution by race in the School Census?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What is the enrollment distribution by race in the School Census?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette shows how to use educabR to explore the racial composition
of school enrollment using the School Census (Censo Escolar) microdata.

```{r setup}
library(educabR)
library(dplyr)
library(ggplot2)
```

## Downloading School Census data

The School Census contains multiple file types. The enrollment data
(matriculas) is the one with student-level information including race.

```{r download}
# List available files to find the enrollment file
list_censo_files(2023)

# Download enrollment data - filter by state to keep manageable
# Full national data is very large (50+ million rows)
censo_sp <- get_censo_escolar(year = 2023, uf = "SP", n_max = 100000)
```

## Race/ethnicity distribution

The School Census uses the `tp_cor_raca` variable with the following
coding from IBGE:

| Code | Category |
|------|----------|
| 0 | Not declared |
| 1 | White (Branca) |
| 2 | Black (Preta) |
| 3 | Mixed race (Parda) |
| 4 | Asian (Amarela) |
| 5 | Indigenous (Indigena) |

```{r race-distribution}
race_labels <- c(
  "0" = "Not declared",
  "1" = "White",
  "2" = "Black",
  "3" = "Mixed race",
  "4" = "Asian",
  "5" = "Indigenous"
)

censo_sp |>
  filter(!is.na(tp_cor_raca)) |>
  mutate(race = race_labels[as.character(tp_cor_raca)]) |>
  count(race, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = reorder(race, pct), y = pct)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Student Enrollment by Race - Sao Paulo (2023)",
    x     = NULL,
    y     = "Percentage (%)"
  ) +
  theme_minimal()
```

## Race distribution by school type

```{r by-school-type}
censo_sp |>
  filter(!is.na(tp_cor_raca), tp_cor_raca != 0) |>
  mutate(
    race = race_labels[as.character(tp_cor_raca)],
    school_type = case_when(
      tp_dependencia == 1 ~ "Federal",
      tp_dependencia == 2 ~ "State",
      tp_dependencia == 3 ~ "Municipal",
      tp_dependencia == 4 ~ "Private"
    )
  ) |>
  count(school_type, race) |>
  mutate(pct = n / sum(n) * 100, .by = school_type) |>
  ggplot(aes(x = school_type, y = pct, fill = race)) +
  geom_col(position = "dodge") +
  labs(
    title = "Racial Composition by School Type - SP (2023)",
    x     = NULL,
    y     = "Percentage (%)",
    fill  = "Race"
  ) +
  theme_minimal()
```

## Comparing states

```{r compare-states}
# Download samples from different states
states <- c("SP", "BA", "RS")

censo_list <- lapply(states, function(uf) {
  get_censo_escolar(year = 2023, uf = uf, n_max = 50000)
})

censo_multi <- bind_rows(censo_list)

censo_multi |>
  filter(!is.na(tp_cor_raca), tp_cor_raca != 0) |>
  mutate(race = race_labels[as.character(tp_cor_raca)]) |>
  count(sg_uf, race) |>
  mutate(pct = n / sum(n) * 100, .by = sg_uf) |>
  ggplot(aes(x = race, y = pct, fill = sg_uf)) +
  geom_col(position = "dodge") +
  labs(
    title = "Racial Composition of Enrollment by State (2023)",
    x     = NULL,
    y     = "Percentage (%)",
    fill  = "State"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
