---
title: "Which municipalities improved the most in IDEB?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Which municipalities improved the most in IDEB?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette shows how to use educabR to identify which municipalities
had the largest IDEB improvement over time, using the historical series
available in the IDEB spreadsheets.

```{r setup}
library(educabR)
library(dplyr)
library(ggplot2)
```

## Downloading IDEB data

Each IDEB file already contains the full historical series (all editions
from 2005 to the current year), so a single download is enough.

```{r download}
ideb <- get_ideb(year = 2023, stage = "anos_iniciais", level = "municipio")
glimpse(ideb)
```

## Calculating IDEB improvement

The columns `vl_observado_YYYY` contain the observed IDEB for each
edition. We compare the earliest and latest available values.

```{r improvement}
ideb_change <- ideb |>
  filter(!is.na(vl_observado_2005), !is.na(vl_observado_2023)) |>
  mutate(
    change     = vl_observado_2023 - vl_observado_2005,
    pct_change = change / vl_observado_2005 * 100
  )
```

## Top 20 municipalities by absolute improvement

```{r top-absolute}
ideb_change |>
  slice_max(change, n = 20) |>
  ggplot(aes(
    x = reorder(paste(no_municipio, sg_uf, sep = " - "), change),
    y = change
  )) +
  geom_col(fill = "#2a9d8f") +
  coord_flip() +
  labs(
    title = "Top 20 Municipalities by IDEB Improvement (2005-2023)",
    subtitle = "Early elementary (anos iniciais)",
    x = NULL,
    y = "IDEB change (2023 - 2005)"
  ) +
  theme_minimal()
```

## Distribution of improvement by state

```{r by-state}
ideb_change |>
  ggplot(aes(x = reorder(sg_uf, change, FUN = median), y = change)) +
  geom_boxplot(fill = "#2a9d8f", alpha = 0.6, outlier.size = 0.5) +
  coord_flip() +
  labs(
    title = "IDEB Improvement Distribution by State (2005-2023)",
    subtitle = "Early elementary (anos iniciais)",
    x = NULL,
    y = "IDEB change"
  ) +
  theme_minimal()
```

## IDEB trajectory for selected municipalities

```{r trajectories}
# Pick the top 5 improvers
top5 <- ideb_change |>
  slice_max(change, n = 5) |>
  pull(co_municipio)

# Reshape to long format for plotting
library(tidyr)

ideb |>
  filter(co_municipio %in% top5) |>
  select(co_municipio, no_municipio, sg_uf, starts_with("vl_observado_")) |>
  pivot_longer(
    cols      = starts_with("vl_observado_"),
    names_to  = "year",
    values_to = "ideb_score"
  ) |>
  mutate(
    year  = as.integer(gsub("vl_observado_", "", year)),
    label = paste(no_municipio, sg_uf, sep = " - ")
  ) |>
  filter(!is.na(ideb_score)) |>
  ggplot(aes(x = year, y = ideb_score, color = label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "IDEB Trajectory - Top 5 Improvers",
    x     = "Year",
    y     = "IDEB",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Municipalities that declined

```{r declined}
ideb_change |>
  filter(change < 0) |>
  slice_min(change, n = 15) |>
  ggplot(aes(
    x = reorder(paste(no_municipio, sg_uf, sep = " - "), change),
    y = change
  )) +
  geom_col(fill = "#e76f51") +
  coord_flip() +
  labs(
    title = "Municipalities with Largest IDEB Decline (2005-2023)",
    subtitle = "Early elementary (anos iniciais)",
    x = NULL,
    y = "IDEB change (2023 - 2005)"
  ) +
  theme_minimal()
```
